---
output:
  pdf_document: default
  html_document: default
---

# STA322 Project 1
## Austin Huang and Abby Li

### Sampling Data

```{r}
library(dplyr)
library(tidyr)
library(survey)

colleges <- read.csv("colleges.csv")
```

```{r}

colleges_q1 <-  colleges %>%
  select(INSTNM, UGDS, CONTROL, CIP27BACHL, MD_EARN_WNE_5YR, TUITIONFEE_IN) %>%
  mutate (type = case_when(
    CONTROL == 1 ~ "Public",
    CONTROL %in% c(2, 3) ~"Private",
    TRUE ~ NA_character_
  )) %>%
  mutate(size = case_when(
    UGDS > 15000 ~ "Large",
    UGDS >= 2000 & UGDS <= 15000 ~ "Medium",
    UGDS < 2000 ~ "Small",
    TRUE ~ NA_character_  
  ))

# creating the six strata 
colleges_q1 <- colleges_q1 %>%
  mutate(stratum = case_when(
    type == "Public" & size == "Large"  ~ "Public-Large",
    type == "Public" & size == "Medium" ~ "Public-Medium",
    type == "Public" & size == "Small"  ~ "Public-Small",
    type == "Private" & size == "Small" ~ "Private-Small",
    type == "Private" & size %in% c("Medium","Large")~ "Private-Medium/Large",
    TRUE ~ NA_character_
  ))

colleges_q1 <- colleges_q1 %>% 
  drop_na()

```

```{r}
# sample with proportional allocation

N_total <- nrow(colleges_q1)
n_total <- 100

set.seed(322)

# ---------- 1) Compute stratum population sizes N_h ----------
stratum_info <- colleges_q1 %>% 
  group_by(stratum) %>% 
  summarise(N_h = n()) %>% 
  ungroup()

# ---------- 2) Proportional allocation: raw n_h ----------
stratum_info <- stratum_info %>% 
  mutate(n_h_raw = n_total * (N_h / sum(N_h)))

# ---------- 3) Round n_h to integers while ensuring sum = n_total ----------
# simple strategy: floor then distribute remainder to strata with largest fractional parts
stratum_info <- stratum_info %>% 
  mutate(n_floor = floor(n_h_raw),
         frac = n_h_raw - n_floor)

remainder <- n_total - sum(stratum_info$n_floor)
if (remainder > 0) {
  # give +1 to the strata with largest fractional parts
  add_idx <- order(stratum_info$frac, decreasing = TRUE)[1:remainder]
  stratum_info$n_final <- stratum_info$n_floor
  stratum_info$n_final[add_idx] <- stratum_info$n_final[add_idx] + 1
} else {
  stratum_info$n_final <- stratum_info$n_floor
}

# If any n_final == 0 (small strata), enforce a minimum of 1:
min_per_stratum <- 1
zeros <- which(stratum_info$n_final < min_per_stratum)
if(length(zeros)>0){
  for(i in zeros) stratum_info$n_final[i] <- min_per_stratum
  # reduce from the largest strata to keep sum == n_total
  while(sum(stratum_info$n_final) > n_total){
    # pick stratum with largest n_final to decrement (but keep >= min)
    idx <- which.max(stratum_info$n_final)
    if(stratum_info$n_final[idx] > min_per_stratum) stratum_info$n_final[idx] <- stratum_info$n_final[idx] - 1
    else break
  }
}
# check
stopifnot(sum(stratum_info$n_final) == n_total)


# ---------- 4) Join n_h back to frame and sample ----------
colleges_df <- colleges_q1 %>% left_join(stratum_info %>% select(stratum, N_h, n_final), by = "stratum")
names(colleges_df)[names(colleges_df) == "n_final"] <- "n_h"

# draw SRS without replacement within each stratum
sampled <- colleges_df %>%
  group_by(stratum) %>%
  sample_n(size = unique(n_h)) %>%
  ungroup()

# ---------- 5) Compute design weights ----------
# For SRS within stratum, inclusion prob pi = n_h / N_h, so weight = 1/pi = N_h / n_h
sampled <- sampled %>%
  mutate(pi = n_h / N_h,
         weight = N_h / n_h)
```

### Question 1
What is an estimate of the total number of undergraduate students enrolled in colleges? Design your survey with the goal of answering this question most accurately. 

```{r}
des <- svydesign(id = ~1, strata = ~stratum, weights = ~weight, fpc = ~N_h, data = sampled)

total_undergrads <- svytotal(~UGDS, des)
total_undergrads
confint(total_undergrads)
```
The total number of undergraduate students enrolled in college is 13,056,855 with a standard error of 1,157,820. We are 95% confident that the true total number of undergraduates is between 10,787,569 and 15,326,140. 

### Question 2
What fraction of colleges have a major in statistical science? Use your sample to estimate this fraction; do not get the answer by filtering the schools. 

### Question 3
What is the average of the median earnings among alumni from public schools? From private schools? (combine for profit and not for profit private schools)

```{r}
avg_median_earnings <- svyby(~MD_EARN_WNE_5YR, ~type, des, svymean, na.rm = TRUE)
avg_median_earnings
confint(avg_median_earnings)
```
The average of the median earnings among alumni 5 years after completion of private colleges is 48267.42 with a standard error of 2303.50. The average of the median earnings among alumni 5 years after completion of public colleges is 49500.21 with a standard error of 1286.48.
The 95% confidence interval for median earnings for private school alum is between 43,752 and 52,782. The 95% confidence interval for median earnings for public school alum is between 46,978 and 52,021. 

### Question 4
What is the average of the tuition for in-state students from public schools? From private schools? (combine for profit and not for profit private schools)

```{r}
avg_median_debt <- svyby(~TUITIONFEE_IN, ~type, des, svymean, na.rm = TRUE)
avg_median_debt
confint(avg_median_debt)
```

The average of tuition for in-state students in private colleges is 28,834 with a standrad error of 2017. The average tuition for in-state students in public colleges is 5,990 with a standard error of 491. 
The 95% CI for average tuition of in-state students in private colleges is between 24880 and 32788. The 95% CI for average tuition of in-state students in public colleges is between 5026 and 6955. 

